Dataflow-Guided Filtering for Efficient and Adjustable Run-Time Monitoring

### 2014-05-30 arch.tex
Minor edits based on Ed's comments.

% Example dataflow operations
\begin{table*}[tb]
  \begin{center}
    \begin{small}
    \input{tabs/dataflow_operations}
    \end{small}
    \caption{Filtering conditions and operations for clean flags.}
    \label{tab:arch.dataflow_operations}
    \vspace{-0.2in}
  \end{center}
\end{table*}

@@

Table~\ref{tab:arch.dataflow_operations} summarizes when when monitoring events
can be filtered out due to clean and invalid metadata. In addition, it shows
what metadata needs to be modified on a filtered or dropped monitoring event.
Note that for simplicity the policies for clean and invalid flags are shown
separately. However, the dataflow engine actually operates on pairs of these
flags using a combined policy.

In terms of filtering using clean flags, array allocations are never filtered
since these will always involve non-clean metadata (i.e., setting new base and
bounds). For ALU operations, if both source operands are clean, then the event
can be filtered by setting the destination register as clean. For load and
store, only a single source needs to be checked to be clean in order for the
event to be filtered. Similarly, the destination register or memory location
needs to be marked as clean if the event is filtered.

The invalid flags for array bounds check operate similarly to the clean flags.
When ALU, load, or store events are dropped or filtered, the destination
register or memory location is marked as invalid. In addition, we need to
handle the possibility of dropping an array allocation. In this case, the array
pointer is marked as invalid. In terms of filtering out invalid events, for an
ALU event, if either of the source registers is invalid then the event can be
filtered. For load and store instructions, if the single source operand is
invalid, then the event can be filtered.

@@

Our approach can be generally applied to a wide range of monitoring schemes. This
is possible because the use of clean and invalidation flags is largely agnostic
to the semantics of the specific monitoring scheme. That is, the dataflow
engine must be configured based on the dataflow of the monitoring scheme, but
it does not need detailed information of what the monitoring scheme does.

We do note some limitations to the proposed approach. Previous work
\cite{fade-hpca14} that has looked at filtering in order to reduce monitoring
overheads has also filtered out redundant updates where the source metadata
matches the destination metadata. Since we do not read the actual metadata, our
architecture is not able to identify these redundant updates. However, by using
only 2-bit flags, the bandwidth need of our dataflow engine is reduced.  
In addition, we note that filtering using clean flags works well when a large
portion of instructions are not relevant to the monitoring scheme (e.g., array
bounds check only cares about array pointers). However, if most instruction
have dirty metadata, then very little filtering will be possible.

@@

Another limitation is that the dataflow engine, as described, only marks one
flag location on each monitoring event. Similarly, the dataflow engine can only
read in two sets of flags in order to determine whether an event can be
filtered or not. This was done because we found that for many monitoring
schemes, the monitoring operations corresponded to program instructions, which
read up to two operands and write to one destination.  For monitoring schemes
which update multiple pieces of metadata on a monitoring event, it may be
adequate to associate a clean and invalid flag with a single piece of metadata.  For
example, a scheme may use a data structure that includes several pieces of
related metadata. On a drop, one of these pieces could be marked as invalid in
order to indicate that the entire structure is invalid.  Similarly, marking one
specified piece as clean may indicate that the entire data structure's metadata
is clean.  Alternatively, the read and write capabilities of the dataflow
engine could be expanded to support more complex monitoring schemes.

>>>>

% Example dataflow operations
\begin{table*}[tb]
  \begin{center}
    \begin{small}
    \input{tabs/dataflow_operations}
    \end{small}
    \caption{Filtering conditions and dropping/filtering operations for null and invalid flags.}
    \label{tab:arch.dataflow_operations}
    \vspace{-0.2in}
  \end{center}
\end{table*}

@@

Table~\ref{tab:arch.dataflow_operations} summarizes when monitoring events
can be filtered out due to null and invalid metadata. In addition, it shows
what metadata needs to be modified on a filtered or dropped monitoring event.
Note that for simplicity the policies for null and invalid flags are shown
separately. However, the dataflow engine actually operates on pairs of these
flags using a combined policy.

In terms of filtering using null flags, array allocations are never filtered
since these will always involve non-null metadata (i.e., setting new base and
bounds). For ALU operations, if both source operands are null, then the event
can be filtered by setting the destination register as null. For load and
store, only a single source needs to be checked to be null in order for the
event to be filtered. Similarly, the destination register or memory location
needs to be marked as null if the event is filtered.

The invalid flags for array bounds check operate similarly to the null flags.
When ALU, load, or store events are dropped or filtered, the destination
register or memory location is marked as invalid. In addition, we need to
handle the possibility of dropping an array allocation. In this case, the array
pointer is marked as invalid. In terms of filtering out invalid events, for an
ALU event, if either of the source registers is invalid then the event can be
filtered. For load and store instructions, if the single source operand is
invalid, then the event can be filtered.

@@

Our approach can be generally applied to a wide range of monitoring schemes. This
is possible because the use of null and invalidation flags is largely agnostic
to the semantics of the specific monitoring scheme. That is, the dataflow
engine must be configured based on the dataflow of the monitoring scheme, but
it does not need detailed information of what the monitoring scheme does.

We do note some limitations to the proposed approach. Previous work
\cite{fade-hpca14} that has looked at filtering in order to reduce monitoring
overheads has also filtered out redundant updates where the source metadata
matches the destination metadata. Since we do not read the actual metadata, our
architecture is not able to identify these redundant updates. However, by using
only 2-bit flags, the bandwidth need of our dataflow engine is reduced.  
In addition, we note that filtering using null flags works well when a large
portion of instructions are not relevant to the monitoring scheme (e.g., array
bounds check only cares about array pointers). However, if most instruction
have non-null metadata, then very little filtering will be possible.

@@

Another limitation is that the dataflow engine, as described, only marks one
flag location on each monitoring event. Similarly, the dataflow engine can only
read in two sets of flags in order to determine whether an event can be
filtered or not. This was done because we found that for many monitoring
schemes, the monitoring operations corresponded to program instructions, which
read up to two operands and write to one destination.  For monitoring schemes
which update multiple pieces of metadata on a monitoring event, it may be
adequate to associate a null and invalid flag with a single piece of metadata.  For
example, a scheme may use a data structure that includes several pieces of
related metadata. On a drop, one of these pieces could be marked as invalid in
order to indicate that the entire structure is invalid.  Similarly, marking one
specified piece as null may indicate that the entire data structure's metadata
is null.  Alternatively, the read and write capabilities of the dataflow
engine could be expanded to support more complex monitoring schemes.

### 2014-05-30 evaluation.tex
Minor edits based on Ed's comments.

In order to explore the generality of the architecture for
different monitors, we implemented three different monitors: uninitialized
memory check (UMC), array bounds check (BC), and dynamic information flow
tracking (DIFT).  Uninitialized memory check seeks to detect loading from
memory locations that are not initialized first.  Array bounds check, as
mentioned in Section~\ref{sec:monitoring} is a monitoring scheme that aims to
detect buffer overflows where memory accesses go beyond the boundaries of an
array. We modify the implementation of {\tt malloc} to set base and bound
metadata information. Dynamic information flow tracking is a security
monitoring scheme,
which detects when information from untrusted sources is used to affect the
program control flow. For our DIFT implementation, we set mark data read from
files as untrusted. This data is marked with a 32-bit metadata identifier so
that if an error is detected, it is possible to have information about where
the information originated from.

@@

Adding the dropping hardware in order to enable adjustable overheads adds
overheads in terms of area and power. We use McPat \cite{mcpat-micro09} to get
a first-order estimate of these area and power overheads in a 40 nm technology
node. McPat estimates the main core area as 1.96 mm$^2$ and the peak power usage as
121 mW averaged across all benchmarks. The average runtime power usage was 38.2 
mW. These area and power numbers consist of the core and
L1 cache, but do not include L2 cache, memory controller, and other
peripherals. The power numbers include dynamic as well as static (leakage)
power. For the dropping hardware, the ALUs, MIRF, MIC, and
configuration tables are modeled using the corresponding objects in McPat. We
note that this is only a rough area and power result since components such as the
wires connecting these modules have not been modeled. However, this gives a
sense of the order-of-magnitude overheads involved with implementing our
approach.

For a 1 kB MIC, an additional 0.183 mm$^2$ of silicon area is needed, an
increase of 9\% of the main core area. Table~\ref{tab:evaluation.area_power}
shows the peak and runtime power overheads. The peak power is 13-19 mW, which is
11-16\% of the main core's peak power usage. The average runtime power is 8-11
mW, corresponding to 21-29\% of the main core's runtime power.

>>>>

In order to explore the generality of the architecture for
different monitors, we implemented three different monitors: uninitialized
memory check (UMC), array bounds check (BC), and dynamic information flow
tracking (DIFT).  Uninitialized memory check seeks to detect loading from
memory locations that are not initialized first.  Array bounds check, as
mentioned in Section~\ref{sec:monitoring} is a monitoring scheme that aims to
detect buffer overflows where memory accesses go beyond the boundaries of an
array. We modify the implementation of {\tt malloc} to set base and bound
metadata information. Dynamic information flow tracking is a security
monitoring scheme,
which detects when information from untrusted sources is used to affect the
program control flow. For the benchmarks we consider, we mark data read from
files as untrusted. We implemented a multi-bit DIFT scheme which marks
untrusted data with a 32-bit metadata identifier so
that if an error is detected, it is possible to have information about where
the data originated from.

@@

Adding the dataflow engine in order to enable filtering and adjustable adjustable adds
overheads in terms of area and power. We use McPat \cite{mcpat-micro09} to get
a first-order estimate of these area and power overheads in a 40 nm technology
node. McPat estimates the main core area as 1.96 mm$^2$ and the peak power usage as
121 mW averaged across all benchmarks. The average runtime power usage was 38.2 
mW. These area and power numbers consist of the core and
L1 cache, but do not include L2 cache, memory controller, and other
peripherals. The power numbers include dynamic as well as static (leakage)
power. For the dataflow engine, we modeled the ALUs used for address
calculation, the dataflow flag register file and cache, the configuration
tables, and the filter decision table. These were modeled using the
corresponding objects in McPat. We
note that this is only a rough area and power result since components such as the
wires connecting these modules have not been modeled. However, this gives a
sense of the order-of-magnitude overheads involved with implementing our
approach.

Our results show that an additional 0.183 mm$^2$ of silicon area is needed, an
increase of 9\% of the main core area. Table~\ref{tab:evaluation.area_power}
shows the peak and runtime power overheads. The peak power is 13-19 mW, which is
11-16\% of the main core's peak power usage. The average runtime power is 8-11
mW, corresponding to 21-29\% of the main core's runtime power.

### 2014-05-30 optimizations.tex
Revise evaluation section.
Minor edits based on Ed's comments.

Although the monitoring core operates in parallel to the main core, the
overheads of programmable monitoring can be high due to the multi-cycle
operation of the monitoring core.
Figure~\ref{dummy} shows the full
monitoring overheads for three example monitoring extensions: array bounds
check (BC), uninitialized memory check (UMC), and dynamic information flow
tracking (DIFT) (Section~\ref{sec:evaluation} details the experimental methodology). In order to reduce these overheads, we present two
optimizations. The first optimization is to filter out monitoring events which relate to
null metadata which do not need to processed by the monitoring core (Section~\ref{sec:optimizations.filter}). Second, in order to allow
further reduction in overheads, we propose to perform partial monitoring. This
allows us to further reduce the overheads of monitoring by trading-off
monitoring coverage. Figure~\ref{fig:optimizations.overview} shows an overview
of
these ideas.

@@

The instruction-grained run-time monitoring schemes we focus on typically
manage a set of metadata information. This metadata information is used to
verify security or reliability properties. For example, for array bounds
check, the monitoring core stores metadata describing the base and bound
addresses for array pointers. This metadata is then checked on load and store
instructions. However, if the metadata for an event is null, then there is no
need to perform the check since this represents a non-array pointer value. By
identifying and filtering out these null monitoring events, we can reduce the
overheads of
monitoring by reducing the amount of work the monitoring core must do.

>>>>

Although the monitoring core operates in parallel to the main core, the
overheads of programmable monitoring can be high due to the multi-cycle
operation of the monitoring core.
 In order to reduce these overheads, we present two
optimizations. The first optimization is to filter out monitoring events which relate to
null metadata which do not need to processed by the monitoring core (Section~\ref{sec:optimizations.filter}). Second, in order to allow
further reduction in overheads, we propose to perform partial monitoring. This
allows us to further reduce the overheads of monitoring by trading-off
monitoring coverage. Figure~\ref{fig:optimizations.overview} shows an overview
of
these ideas.

@@

The instruction-grained run-time monitoring schemes that we focus on typically
manage a set of metadata information. This metadata information is used to
verify security or reliability properties. For example, for array bounds
check, the monitoring core stores metadata describing the base and bound
addresses for array pointers. This metadata is then checked on load and store
instructions. However, if the metadata for an event is null, then there is no
need to perform the check since this represents a non-array pointer value. By
identifying and filtering out these null monitoring events, we can reduce the
mount of work done by the monitoring core and thus reduce the overheads of
monitoring.

### 2014-05-30 policies.tex
Minor edits based on Ed's comments.

\begin{figure}
  \begin{center}
  \subfloat[unrestricted dropping]{
    \includegraphics[width=\columnwidth]{figs/all_drop.pdf}
    \label{fig:policies.all_drop}
  }
  \vspace{-0.1in}
  \\
  \subfloat[source-only dropping]{
    \includegraphics[width=\columnwidth]{figs/source_drop.pdf}
    \label{fig:policies.source_drop}
  }
  \vspace{-0.1in}
  \\
  \subfloat[split-point dropping]{
    \includegraphics[width=\columnwidth]{figs/min_drop.pdf}
    \label{fig:policies.min_drop}
  }
  \end{center}
  \vspace{-0.2in}
  \caption{Comparison of dropping policies. Blue (dark) nodes indicate where
  dropping decisions are made.}
  \label{fig:policies.policies}
\end{figure}

@@

Another possible method is to make the dropping decisions somewhere in between
source-only dropping and unrestricted dropping. From the graph in 
Figure~\ref{fig:policies.dataflow_graph}, we can see that if the dropping
decision is made at event {\tt C} then we can skip event {\tt E} with no wasted
work. If we instead choose to perform monitoring for event {\tt C}, then we want to perform all
monitoring on that flow. Similarly, event {\tt F} creates a decision point for
the bottom flow that will result in no wasted work and dropping event {\tt A}
allows the entire shown metadata flow to be dropped with no wasted work. If the
program can be
analyzed or profiled to identify these instructions which, if dropped, lead to
no wasted work, then this information can be used at run-time to minimize the
amount of wasted work. We refer to this policy as \emph{sub-flow dropping}.
Figure~\ref{fig:policies.min_drop} shows the points were dropping decisions are
made for this policy.
Min-event dropping is a coarser-grained decision than unrestricted dropping, but
finer-grained that source dropping. Thus, its ability to match overheads is
expected to sit between these other two policies. Similarly, there still exists
edge cases where wasted work can occur, but typically we expect much less
wasted work than unrestricted 
dropping. Thus, we expect the coverage achieved by split-point dropping for a
certain overhead to fall between source and unrestricted dropping. These three
dropping policies create a trade-off space between coverage achieved and
ability to match an overhead target.
Figure~\ref{fig:policies.trade_off} summarizes this trade-off between coverage
and ability to match overhead.

>>>>

\begin{figure}
  \begin{center}
  \subfloat[unrestricted dropping]{
    \includegraphics[width=\columnwidth]{figs/all_drop.pdf}
    \label{fig:policies.all_drop}
  }
  \vspace{-0.1in}
  \\
  \subfloat[source-only dropping]{
    \includegraphics[width=\columnwidth]{figs/source_drop.pdf}
    \label{fig:policies.source_drop}
  }
  \vspace{-0.1in}
  \\
  \subfloat[sub-flow dropping]{
    \includegraphics[width=\columnwidth]{figs/min_drop.pdf}
    \label{fig:policies.min_drop}
  }
  \end{center}
  \vspace{-0.2in}
  \caption{Comparison of dropping policies. Blue (dark) nodes indicate where
  dropping decisions are made.}
  \label{fig:policies.policies}
\end{figure}

@@

Another possible method is to make the dropping decisions somewhere in between
source-only dropping and unrestricted dropping. From the graph in 
Figure~\ref{fig:policies.dataflow_graph}, we can see that if the dropping
decision is made at event {\tt C} then we can skip event {\tt E} with no wasted
work. If we instead choose to perform monitoring for event {\tt C}, then we want to perform all
monitoring on that flow. Similarly, event {\tt F} creates a decision point for
the bottom flow that will result in no wasted work and dropping event {\tt A}
allows the entire shown metadata flow to be dropped with no wasted work. If the
program can be
analyzed or profiled to identify these instructions which, if dropped, lead to
no wasted work, then this information can be used at run-time to minimize the
amount of wasted work. We refer to this policy as \emph{sub-flow dropping}.
Figure~\ref{fig:policies.min_drop} shows the points were dropping decisions are
made for this policy.
Sub-flow dropping is a coarser-grained decision than unrestricted dropping, but
finer-grained that source dropping. Thus, its ability to match overheads is
expected to sit between these other two policies. Similarly, there still exists
edge cases where wasted work can occur, but typically we expect much less
wasted work than unrestricted 
dropping. Thus, we expect the coverage achieved by sub-flow dropping for a
certain overhead to fall between source and unrestricted dropping. These three
dropping policies create a trade-off space between coverage achieved and
ability to match an overhead target.
Figure~\ref{fig:policies.trade_off} summarizes this trade-off between coverage
and ability to match overhead.

### 2014-05-30 related.tex
Minor edits based on Ed's comments.

% Filtering
FADE \cite{fade-hpca14} presents a general filtering engine in order to reduce
the overheads of monitoring. It filters out redundant updates and clean checks
by reading metadata values. We do not handle filtering redundant updates, but
we do filter out clean checks as well as clean (i.e., uninitialized)
propagations.  We are able to do this because the dataflow engine can also
propagate the clean information.  In addition, our dataflow engine only
requires reading and writing 2-bit flags rather than the entire metadata.  As
we have shown, our dataflow engine also enables adjustable overheads without
false positives which is not a capability described by FADE.

@@

% Limited Monitoring
In terms of controlling overheads, the Quality Virtual Machine (QVM) is a
modification of the Java Virtual Machine (JVM) that supports run-time
monitoring with controllable overheads \cite{qvm-oopsla08}. Similarly, Huang et
al. created a framework for controlling the overheads of software-based
monitoring \cite{huang-sttt12}. These projects are targeting a similar problem
to the one we address in this paper. However, both projects are focused on
software-based monitoring and thus their mechanisms modify the software
monitoring framework in order to enable or disable monitoring. Instead, our
architecture provides a general hardware mechanism to control overheads of
parallel monitoring architectures. In addition, these works prevented
false positives by enabling and disabling monitoring using a source-dropping
policy. Lo et al. \cite{lo-rtas14} developed a hardware architecture to limit
monitoring in hard real-time systems. Their architecture performs unrestricted 
dropping in order to guarantee strict deadlines. Our system allows for a
trade-off between choosing source-only dropping, unrestricted dropping, or split-point 
dropping depending on the needs of the designer or user which has not been
explored by these previous works.

>>>>

% Filtering
FADE \cite{fade-hpca14} presents a general filtering engine in order to reduce
the overheads of monitoring. It filters out redundant updates and null checks
by reading metadata values. We do not handle filtering redundant updates, but
we do filter out null checks as well as null (i.e., uninitialized)
propagations.  We are able to do this because the dataflow engine can also
propagate the null information.  In addition, our dataflow engine only
requires reading and writing 2-bit flags rather than the entire metadata.  As
we have shown, our dataflow engine also enables adjustable overheads without
false positives which is not a capability described by FADE.

@@

% Limited Monitoring
In terms of controlling overheads, the Quality Virtual Machine (QVM) is a
modification of the Java Virtual Machine (JVM) that supports run-time
monitoring with controllable overheads \cite{qvm-oopsla08}. Similarly, Huang et
al. created a framework for controlling the overheads of software-based
monitoring \cite{huang-sttt12}. These projects are targeting a similar problem
to the one we address in this paper. However, both projects are focused on
software-based monitoring and thus their mechanisms modify the software
monitoring framework in order to enable or disable monitoring. Instead, our
architecture provides a general hardware mechanism to control overheads of
parallel monitoring architectures. In addition, these works prevented
false positives by enabling and disabling monitoring using only a source-dropping
policy. Lo et al. \cite{lo-rtas14} developed a hardware architecture to limit
monitoring in hard real-time systems. Their architecture performs unrestricted 
dropping in order to guarantee strict deadlines. Our system allows for a
trade-off between choosing source-only dropping, unrestricted dropping, or sub-flow 
dropping depending on the needs of the designer or user which has not been
explored by these previous works.

### 2014-05-30 tabs/dataflow_operations.tex
Minor edits based on Ed's comments.

% Operations for handling clean flags

@@

\multirow{4}{*}{Clean flag}  
& Array allocation & Never & - \\ \cline{2-4}
& ALU   & Both source registers are clean & Mark destination register as clean \\ \cline{2-4}
& Load  & Source memory location is clean & Mark destination register as clean \\ \cline{2-4}
& Store & Source register is clean & Mark destination memory location as clean \\
\hline\hline

@@

% \multirow{2}{*}{UMC}  
% & Store & Never & - \\ \cline{2-4}
% & Load & Source memory location is dirty & Do nothing \\ 
% \hline\hline
% 
% \multirow{5}{*}{DIFT}  
% & Software set tag & Never & - \\ \cline{2-4}
% & ALU   & Both source registers are clean & Mark destination register as clean \\ \cline{2-4}
% & Load  & Source memory location is clean & Mark destination register as clean \\ \cline{2-4}
% & Store & Source register is clean & Mark destination memory location as clean \\ \cline{2-4}
% & Indirect jump & Source register is clean & Do nothing \\ 
% \hline

>>>>

% Operations for handling null flags

@@

\multirow{4}{*}{Clean flag}  
& Array allocation & Never & - \\ \cline{2-4}
& ALU   & Both source registers are null & Mark destination register as null \\ \cline{2-4}
& Load  & Source memory location is null & Mark destination register as null \\ \cline{2-4}
& Store & Source register is null & Mark destination memory location as null \\
\hline\hline

@@

% \multirow{2}{*}{UMC}  
% & Store & Never & - \\ \cline{2-4}
% & Load & Source memory location is dirty & Do nothing \\ 
% \hline\hline
% 
% \multirow{5}{*}{DIFT}  
% & Software set tag & Never & - \\ \cline{2-4}
% & ALU   & Both source registers are null & Mark destination register as null \\ \cline{2-4}
% & Load  & Source memory location is null & Mark destination register as null \\ \cline{2-4}
% & Store & Source register is null & Mark destination memory location as null \\ \cline{2-4}
% & Indirect jump & Source register is null & Do nothing \\ 
% \hline
